---
title: "LuValle Independent Study, Chaotic Modeling"
author: "Luke Beebe"
date: "2024-05-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(lars)
```

Professor LuValle touted his method of predicting future weather events using the method of *multi-view embedding* which ordinarily *embeds* (integrates) information from *multiple views* (data sources) into a common low-dimensional space. While this approach is commonly used when dealing with data sets that contain information from different modalities (text, images, structured data...etc.) it is used in LuValle's code to inspect different variables at different lags (moments) of time commonly associated with time-series models. It samples random assortments of variables and lags and runs a least-angle regression using the most similar trend in data according to a k-nearest neighbors calculation. This is done a specified number of times to give us a multimodal distribution of predictions representing all different combinations of variables.

We were able to rewrite the code, add features, and compare this model to another. We rewrote the code as a means to understand and clean up the already existing code. We added a feature that allows you to extract the *most predictive* variables within a specified range for each predicted season. We then compared this model to an ARIMA model to test their advantages.

The following explains the above paragraph in more detail, showing how the code works by guiding us towards a prediction.

### Data

The initial data is a .csv file containing *year, month, Hawaii Precipitation, Hawaii Temperature Average, MEI, PDO, IOD, AO1, PDO1, Sunspot measurements, Fresno Precipitation, Fresno Temperature Average*.
```{r data1}
weather <- read.csv("/Users/lukebeebe/Documents/School/Rutgers/2024 Spring/Independent Study/HF_data.csv")
head(weather)
```
We need to add seasons to the data, then average the quantities 
```{r data2}
season <- c(4,4,1,1,1,2,2,2,3,3,3,4)
weather <- cbind(season, weather)
```


### Code (recoded)

### Most Used Predictive Variables

### Comparison with ARIMA Time-Series


